---
title: "HR association between CD-PRS and colitis adverse event in GeRI cohort"
output: html_notebook
Author: "Pooja Middha"
Date: "Feb 15, 2023"
---

Aim: Run Cox proportional hazard analysis between CD-PRS and GI adverse event phenotype in GeRI


```{r}
# Load Library
library(data.table)
library(dplyr)
library(ggplot2)
library(cowplot)
library(pROC)
library(ggrepel)
library(survival)
library(ggpubr)

options(stringsAsFactors = F)
options(scipen = 999)

# Project directory
proj_dir <- "~/OneDrive - University of California, San Francisco/Proj_GeRI_GI-irAEs/"
proj_dir
```

Load the GeRI phenotype data and PRS data

```{r}
# Pheno data
pheno <- fread(paste0(proj_dir, 
                      "Data/combined_allsites_colitis_pheno_data_with_colitis_severity_2023-02-15.txt"))
pheno <- pheno[,-1]

# PRS data 
prs <- fread(paste0(proj_dir, "Results/CD/geri_cd_prs.txt"))

# PC data
pc <- fread(paste0(proj_dir, "Data/pca_allchr_pruned.eigenvec"))
pc <- pc[,-1]

# Combine pheno and pc data
geri_pheno_pc <- left_join(pheno, pc, by = "IID") 
geri_prs_pc <- left_join(geri_pheno_pc, prs, by = "IID")
```

Cox proportional hazards model of CD-PRS and GI-irAE (UNADJUSTED MODEL)

```{r}
# Cox proportional model (UNADJUSTED)
unadj_cox <- coxph(Surv(Time_to_colitis, Colitis) ~ prs, data = geri_prs_pc)
summary(unadj_cox)
unadj_cox_df <- data.frame(exp(cbind(HR = coef(unadj_cox), confint(unadj_cox))))
unadj_cox_df$p <- round(coef(summary(unadj_cox))[,5], digits = 5)
unadj_cox_df$var <- rownames(unadj_cox_df)
colnames(unadj_cox_df) <- c("HR", "LCI", "UCI", "p", "var")

# Save unadjusted cox model results 
write.csv(unadj_cox_df, file = paste0(proj_dir, "Results/CD/unadjusted_PRS_cox_regression.csv"), 
          row.names = F)
```
Cox proportional hazards model of CD-PRS and GI-irAE (ADJUSTED MODEL - AGE, BIOLOGICAL GENDER, PCs)

```{r}
# Cox proportional model (ADJUSTED)
adj_cox <- coxph(Surv(Time_to_colitis, Colitis) ~ prs + Age + female + histo_ADC + histo_SQCLC + msk + 
                   ucsf + pmcc + combination_therapy + PC1 + PC2 + PC3 + PC4 + PC5, data = geri_prs_pc)
summary(adj_cox)
adj_cox_df <- data.frame(exp(cbind(HR = coef(adj_cox), confint(adj_cox))))
adj_cox_df$p <- round(coef(summary(adj_cox))[,5], digits = 5)
adj_cox_df$var <- rownames(adj_cox_df)
colnames(adj_cox_df) <- c("HR", "LCI", "UCI", "p","var")

# Save unadjusted cox model results 
write.csv(adj_cox_df, file = paste0(proj_dir, "Results/CD/adjusted_PRS_cox_regression.csv"), 
          row.names = F)
```

Plot adjusted and unadjusted Cox model together in one figure

```{r}
# Unadjusted and adjusted model plots
unadj_hr <- unadj_cox_df
unadj_hr$model <- "UNADJ"
adj_hr <- adj_cox_df[1,]
adj_hr$model <- "ADJ"

# Rbind above two data frames
comb_hr <- bind_rows(unadj_hr, adj_hr)

# Plot both ORs side by side
hr_p <- ggplot(comb_hr, aes(x=HR, y=model)) + 
  geom_errorbarh(aes(xmin=LCI, xmax=UCI), size=1, height=.1, color="#E69F00") + 
  geom_point(size=2, color="#0072B2") + 
  scale_y_discrete(labels = c("Adjusted", "Unadjusted")) +
  theme_bw() + 
  theme(panel.grid = element_blank()) + 
  labs(x="Model", y="HR (95% CI)") + 
  geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed") 

hr_p

# Export the plot
jpeg(paste0(proj_dir, "Results/CD/HRplot_adjusted_unadjusted_cox_regression.jpeg"), width=4450, height=(2800), res=350)
hr_p
dev.off()
```

# STANDARDIZED CD-PRS AND RELATED ANALYSIS
Standardize PRS by unit standard deviation of controls

```{r}
# Standard deviation of CD-PRS in contorls
#sd_con_prs <- sd(gi_control$prs)

# Divide PRS of all individuals by SD of PRS from controls
#geri_prs_pc$sPRS <- as.numeric(geri_prs_pc$prs/sd_con_prs)
#gi_case$sPRS <- as.numeric(gi_case$prs/sd_con_prs)
#gi_control$sPRS <- as.numeric(gi_control$prs/sd_con_prs)
```

Standardize with mean 0 and standard deviation 1

```{r}
# Separate GI adverse event cases and controls
gi_case <- subset(geri_prs_pc, geri_prs_pc$Colitis == 1)
gi_control <- subset(geri_prs_pc, geri_prs_pc$Colitis == 0)

# Compute mean PRS from RAW PRS for cases and controls
case_mean_prs <- gi_case %>% pull(prs) %>% mean() %>% signif(6)
control_mean_prs <- gi_control %>% pull(prs) %>% mean() %>% signif(6)

# New standardized PRS
geri_prs_pc$sPRS <- as.numeric((geri_prs_pc$prs - mean(gi_control$prs))/sd(gi_control$prs))
```

Cox proportional hazards model of sCD-PRS and GI-irAE (UNADJUSTED MODEL)

```{r}
# Cox proportional model (UNADJUSTED)
sunadj_cox <- coxph(Surv(Time_to_colitis, Colitis) ~ sPRS, data = geri_prs_pc)
summary(sunadj_cox)
sunadj_cox_df <- data.frame(exp(cbind(HR = coef(sunadj_cox), confint(sunadj_cox))))
sunadj_cox_df$p <- round(coef(summary(sunadj_cox))[,5], digits = 5)
sunadj_cox_df$var <- rownames(sunadj_cox_df)
colnames(sunadj_cox_df) <- c("HR", "LCI", "UCI", "p", "var")

# Save unadjusted cox model results 
write.csv(sunadj_cox_df, file = paste0(proj_dir, "Results/CD/unadjusted_sPRS_cox_regression.csv"), 
          row.names = F)
```
Cox proportional hazards model of sCD-PRS and GI-irAE (ADJUSTED MODEL - AGE, BIOLOGICAL GENDER, PCs)

```{r}
# Cox proportional model (ADJUSTED)
sadj_cox <- coxph(Surv(Time_to_colitis, Colitis) ~ sPRS + Age + female + histo_ADC + histo_SQCLC + msk + 
                   ucsf + pmcc + combination_therapy + PC1 + PC2 + PC3 + PC4 + PC5, data = geri_prs_pc)
summary(sadj_cox)
sadj_cox_df <- data.frame(exp(cbind(HR = coef(sadj_cox), confint(sadj_cox))))
sadj_cox_df$p <- round(coef(summary(sadj_cox))[,5], digits = 5)
sadj_cox_df$var <- rownames(sadj_cox_df)
colnames(sadj_cox_df) <- c("HR", "LCI", "UCI", "p","var")

# Save unadjusted cox model results 
write.csv(sadj_cox_df, file = paste0(proj_dir, "Results/CD/adjusted_sPRS_cox_regression.csv"), 
          row.names = F)
```

Plot adjusted and unadjusted Cox model together in one figure

```{r}
# Unadjusted and adjusted model plots
sunadj_hr <- sunadj_cox_df
sunadj_hr$model <- "UNADJ"
sadj_hr <- sadj_cox_df[1,]
sadj_hr$model <- "ADJ"

# Rbind above two data frames
scomb_hr <- bind_rows(sunadj_hr, sadj_hr)

# Plot both ORs side by side
shr_p <- ggplot(scomb_hr, aes(x=HR, y=model)) + 
  geom_errorbarh(aes(xmin=LCI, xmax=UCI), size=1, height=.1, color="#E69F00") + 
  geom_point(size=2, color="#0072B2") + 
  geom_label_repel(aes(label=round(HR, digits = 2)), colour="black", segment.colour="black") + 
  scale_y_discrete(labels = c("Adjusted", "Unadjusted")) +
  theme_bw() + 
  theme(panel.grid = element_blank()) + 
  labs(x="Model", y="HR (95% CI)") + 
  geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed") 

shr_p

# Export the plot
jpeg(paste0(proj_dir, "Results/CD/HRplot_adjusted_unadjusted_cox_regression_standardized_PRS.jpeg"), width=4450, height=(2800), res=350)
shr_p
dev.off()
```

Plot HRs from non-standardized and standardized PRS in one plot from both models

```{r}
# Rbind results from all 4 models (standardized and non-standardized unadjusted and adjusted models)
sall_hr <- bind_rows(comb_hr, scomb_hr)
sall_hr$var[1:2] <- "PRS"

# Plot both ORs side by side
all_hr_p <- ggplot(sall_hr, aes(x=HR, y=model)) + 
             geom_errorbarh(aes(xmin=LCI, xmax=UCI), size=1, height=.1, color="#E69F00") + 
             geom_point(size=2, color="#0072B2") + 
             geom_label_repel(aes(label=round(HR, digits = 2)), colour="black", segment.colour="black") + 
             scale_y_discrete(labels = c("Adjusted", "Unadjusted")) +
             facet_wrap(~var) + 
             theme_bw() + 
             theme(panel.grid = element_blank()) + 
             labs(x="HR (95% CI)", y="Model") + 
             geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed") 

all_hr_p

# Export the plot
jpeg(paste0(proj_dir, "Results/CD/HRplot_adjusted_unadjusted_standardized_nonstandardized_PRS.jpeg"), width = 4450, height = 2800, res = 350)
all_hr_p
dev.off()
```