---
title: "Association between CD-PRS and CD phenotype in UK Biobank"
output: html_notebook
Author: "Pooja Middha"
Date: "Oct 12, 2022"
---

Aim: THis cript has follwing aims: 
1) Check the distribution of CD-PRS in CD cases and controls in UKBB
2) Compare mean CD-PRS between CD cases and controls
3) Run association analysis between CD-PRS and phenotype in UKBB 

```{r}
# Load Library
library(data.table)
library(dplyr)
library(ggplot2)
library(cowplot)
library(pROC)
library(ggrepel)
library(ggpubr)

options(stringsAsFactors = F)
```

Load CD PRS scores calculated by Plink for individuals 

```{r}
# PRS data 
uk_cd_prs <- fread("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/cd_ukb_prs.sscore", header = T)
colnames(uk_cd_prs)[1] <- "FID"
```

Load CD pheno data 

```{r}
# Pheno data
load("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Data/pheno/gi_pheno_final_10112022.Rdata")

# Subset phenotype data to those with PRS score
uk_cd_phen <- subset(final, final$FID %in% uk_cd_prs$FID)
```

Merge genetic and phenotype data

```{r}
# Merge genetic and phenotype data
ukb_cd <- merge(uk_cd_phen, uk_cd_prs, by.x = c("FID", "IID"), by.y = c("FID" ,"IID"))
```

Plot distribution of PRS in cases and controls

```{r}
# Separate CD cases and controls
cd_case <- subset(ukb_cd, ukb_cd$CD == 1)
cd_control <- subset(ukb_cd, ukb_cd$CD == 0)

# Compute mean PRS for cases and controls
case_mean_prs <- cd_case %>% pull(SCORE1_SUM) %>% mean() %>% signif(6)
control_mean_prs <- cd_control %>% pull(SCORE1_SUM) %>% mean() %>% signif(6)

# Plot PRS distribution in cases and controls in UKBB 
a <- ggplot(ukb_cd, aes(x=SCORE1_SUM, fill = as.factor(CD), color = as.factor(CD))) +
            geom_density() + 
            scale_fill_manual(labels=c("Control", "Case"), values=alpha(c("#00AFBB", "#E7B800"),0.3)) + 
            scale_color_manual(labels=c("Control", "Case"), values=alpha(c("#00AFBB", "#E7B800"),0.3)) + 
            theme(panel.grid = element_blank(), 
                  panel.border = element_blank(), 
                  plot.margin = margin(0.2, 0.2, 0.1, 0.2, "cm"), # top, right, bottom, left 
                  legend.position = "bottom") + 
            guides(fill=guide_legend(title="Case-control status")) +
            geom_vline(xintercept = case_mean_prs, linetype = "dashed", color = "red") +
            geom_text(aes(x=case_mean_prs, label=round(case_mean_prs, digits = 2), y=0.1), 
                      colour="red", angle = 0, hjust = -0.15) +
            geom_vline(xintercept = control_mean_prs, linetype = "dashed", color = "blue") + 
            geom_text(aes(x=control_mean_prs, label=round(control_mean_prs, digits = 2), y=0.1), 
                      colour="blue", angle = 0, hjust = 1.2) +
            labs(title = "CD-PRS distribution in UKBB", 
                 x = "CD-PRS",
                 y = "Density", 
                 color = "Case-control status") + 
            scale_y_continuous(expand = expansion (add = 0.004))
a

# Plot mean PRS and confidence intervals and t.test to test differences in means
b <- ggviolin(ukb_cd, x = "CD", y = "SCORE1_SUM", fill = "CD",
              ylab = "CD-PRS", xlab = "Case-Control status", 
              add = "boxplot", add.params = list(fill = "white")) + 
              scale_fill_manual(labels = c("Control", "Case"), values=c("#00AFBB", "#E7B800"))+ 
    stat_compare_means(method = "t.test", label = "p.signif") + 
    stat_compare_means(method = "t.test", label.y = 6.5) + 
    labs(title = "Comparison of mean PRS") + 
    theme(legend.position = "none")
b

# Arrange the two plots in a row 
a2 <- a + theme(legend.position = "none")
c <- plot_grid(a2, b)

# Extract the legend from one of the plots
legend <- get_legend(a + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom"))


# add the legend to the row we made earlier. Give it one-third of 
# the width of one plot (via rel_widths).
d <- plot_grid(c, legend, ncol = 1, rel_heights = c(1, .1))
d

# Export the plot 
jpeg("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/CD_PRS_distribution_and_meanPRS_plot.jpeg", width=4450, height=(2800), res=350)
d
dev.off()
```

Association analysis of CD-PRS and CD phenotype (UNADJUSTED MODEL)

```{r}
# Logistic regression (UNADJUSTED)
unadj <- glm(CD ~ SCORE1_SUM, data = ukb_cd, family = "binomial")
summary(unadj)
unadj_df <- NULL
unadj_df <- data.frame(exp(cbind(OR = coef(unadj), confint(unadj))))
unadj_df$p <- coef(summary(unadj))[,4]
unadj_df$var <- row.names(unadj_df)
colnames(unadj_df) <- c("OR", "LCI", "UCI", "p", "var")

# Save unadjusted model results 
write.table(unadj_df, file = "~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/unadjustedprs_logistic_regression.txt", 
            sep = "\t", quote = F, row.names = F)
```

Association analysis of CD-PRS and CD phenotype (ADJUSTED MODEL - AGE, BIOLOGICAL GENDER, PCs, GENOTYPING ARRAY)

```{r}
# Logistic regression (UNADJUSTED)
adj <- glm(CD ~ SCORE1_SUM + CD_agedx_enrol + sex_self + genotyping.array + 
                PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, 
                data = ukb_cd, family = "binomial")
summary(adj)
adj_df <- NULL
adj_df <- data.frame(exp(cbind(OR = coef(adj), confint(adj))))
adj_df$p <- coef(summary(adj))[,4]
adj_df$var <- row.names(adj_df)
colnames(adj_df) <- c("OR", "LCI", "UCI", "p", "var")

# Save the adjusted model results
write.table(adj_df, file = "~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/adjustedprs_logistic_regression.txt", 
            sep = "\t", quote = F, row.names = F)
```

Plot adjusted and unadjusted model together in one figure

```{r}
# Unadjusted and adjusted model plots
unadj_or <- unadj_df[-1,]
unadj_or$model <- "UNADJ"
adj_or <- adj_df[2,]
adj_or$model <- "ADJ"

# Rbind above two data frames
comb_or <- bind_rows(unadj_or, adj_or)

# Plot both ORs side by side
or_p <- ggplot(comb_or, aes(x=OR, y=model)) + 
             geom_errorbarh(aes(xmin=LCI, xmax=UCI), size=1, height=.1, color="#E7B800") + 
             geom_point(size=2, color="#00AFBB") + 
             scale_y_discrete(labels = c("Adjusted", "Unadjusted")) +
             theme_bw() + 
             theme(panel.grid = element_blank()) + 
             labs(x="Model", y="OR (95% CI)") + 
             geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed") 

or_p

# Export the plot
jpeg("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/plot_adjusted_unadjusted_regression_OR.jpeg")
or_p
dev.off()
```

```{r}
# Plot AUROC of actual PRS in cases and controls together
jpeg("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/auc_cdprs_ukbb.jpeg", res = (800), width=(4450), height = (4450))
rocobj <- plot.roc(ukb_cd$CD, ukb_cd$SCORE1_SUM,
                   main = "AUC-CD", 
                   percent=FALSE,
                   ci = TRUE,                  # compute AUC (of AUC by default)
                   print.auc = TRUE,
                   identity.col="darkgrey", print.auc.cex=par("cex"))
dev.off()
```

# STANDARDIZED CD-PRS AND RELATED ANALYSIS
Standardize PRS by unit standard deviation of controls

```{r}
# Standard deviation of CD-PRS in contorls
#sd_cont <- as.numeric(sd(cd_control$SCORE1_SUM)) # 1.194902

# Divide PRS of all individuals by SD of PRS from controls
#ukb_cd$sPRS <- as.numeric(ukb_cd$SCORE1_SUM/sd_cont)
#cd_case$sPRS <- as.numeric(cd_case$SCORE1_SUM/sd_cont)
#cd_control$sPRS <- as.numeric(cd_control$SCORE1_SUM/sd_cont)
```

Standardize where is it centered around 0

```{r}
# New standardized PRS
ukb_cd$sPRS <- as.numeric((ukb_cd$SCORE1_SUM - mean(cd_control$SCORE1_SUM))/sd(cd_control$SCORE1_SUM))
```

Plot distribution of sPRS in cases and controls

```{r}
# Compute mean sPRS for cases and controls
#case_mean_sprs <- cd_case %>% pull(sPRS) %>% mean() %>% signif(6)
#control_mean_sprs <- cd_control %>% pull(sPRS) %>% mean() %>% signif(6)

# Plot PRS distribution in cases and controls in UKBB 
sa <- ggplot(ukb_cd, aes(x=sPRS, fill = as.factor(CD), color = as.factor(CD))) +
            geom_density() + 
            scale_fill_manual(labels=c("Control", "Case"), values=alpha(c("#CC33FF", "#FFCC00"),0.3)) + 
            scale_color_manual(labels=c("Control", "Case"), values=alpha(c("#CC33FF", "#FFCC00"),0.3)) + 
            theme(panel.grid = element_blank(), 
                  panel.border = element_blank(), 
                  plot.margin = margin(0.2, 0.2, 0.1, 0.2, "cm"), # top, right, bottom, left 
                  legend.position = "bottom") + 
            guides(fill=guide_legend(title="Case-control status")) +
#            geom_vline(xintercept = case_mean_sprs, linetype = "dashed", color = "red") +
#            geom_text(aes(x=case_mean_sprs, label=round(case_mean_sprs, digits = 2), y=0.1), 
#                      colour="red", angle = 0, hjust = -0.15) +
 #           geom_vline(xintercept = control_mean_sprs, linetype = "dashed", color = "blue") + 
#            geom_text(aes(x=control_mean_sprs, label=round(control_mean_sprs, digits = 2), y=0.1), 
#                      colour="blue", angle = 0, hjust = 1.2) +
            labs(title = "sCD-PRS distribution in UKBB", 
                 x = "sCD-PRS",
                 y = "Density", 
                 color = "Case-control status") + 
            scale_y_continuous(expand = expansion (add = 0.004))
sa

# Plot mean PRS and confidence intervals and t.test to test differences in means
sb <- ggviolin(ukb_cd, x = "CD", y = "sPRS", fill = "CD",
              ylab = "sCD-PRS", xlab = "Case-Control status", 
              add = "boxplot", add.params = list(fill = "white")) + 
              scale_fill_manual(labels = c("Control", "Case"), values=c("#CC33FF", "#FFCC00"))+ 
    stat_compare_means(method = "t.test", label = "p.signif") + 
    stat_compare_means(method = "t.test", label.y = 6.5) + 
    labs(title = "Comparison of mean sPRS") + 
    theme(legend.position = "none")
sb

# Arrange the two plots in a row 
sa2 <- sa + theme(legend.position = "none")
sc <- plot_grid(sa2, sb)

# Extract the legend from one of the plots
slegend <- get_legend(sa + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom"))

# add the legend to the row we made earlier. Give it one-third of 
# the width of one plot (via rel_widths).
sd <- plot_grid(sc, slegend, ncol = 1, rel_heights = c(1, .1))
sd

# Export the plot 
jpeg("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/CD_standardized_PRS_distribution_and_meanPRS_plot.jpeg", width=4450, height=(2800), res=350)
sd
dev.off()
```

Association analysis of sCD-PRS and CD phenotype (UNADJUSTED MODEL)

```{r}
# Logistic regression (UNADJUSTED)
sunadj <- glm(CD ~ sPRS, data = ukb_cd, family = "binomial")
summary(sunadj)
sunadj_df <- NULL
sunadj_df <- data.frame(exp(cbind(OR = coef(sunadj), confint(sunadj))))
sunadj_df$p <- coef(summary(sunadj))[,4]
sunadj_df$var <- row.names(sunadj_df)
colnames(sunadj_df) <- c("OR", "LCI", "UCI", "p","var")

# Save unadjusted model results 
write.table(sunadj_df, file = "~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/unadjusted_standardizedprs_logistic_regression.txt", sep = "\t", quote = F, row.names = F)
```

Association analysis of sCD-PRS and CD phenotype (ADJUSTED MODEL - AGE, BIOLOGICAL GENDER, PCs, GENOTYPING ARRAY)

```{r}
# Logistic regression (UNADJUSTED)
sadj <- glm(CD ~ sPRS + CD_agedx_enrol + sex_self + genotyping.array + PC1 + 
                 PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, 
                 data = ukb_cd, family = "binomial")
summary(sadj)
sadj_df <- NULL
sadj_df <- data.frame(exp(cbind(OR = coef(sadj), confint(sadj))))
sadj_df$p <- coef(summary(sadj))[,4]
sadj_df$var <- row.names(sadj_df)
colnames(sadj_df) <- c("OR", "LCI", "UCI", "p","var")

# Save the adjusted model results
write.table(sadj_df, file = "~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/adjusted_standardizedprs_logistic_regression.txt", sep = "\t", quote = F, row.names = F)
```

Plot adjusted and unadjusted model together in one figure

```{r}
# Unadjusted and adjusted model plots
sunadj_or <- sunadj_df[-1,]
sunadj_or$model <- "UNADJ"
sadj_or <- sadj_df[2,]
sadj_or$model <- "ADJ"

# Rbind above two data frames
scomb_or <- bind_rows(sunadj_or, sadj_or)

# Plot both ORs side by side
sor_p <- ggplot(scomb_or, aes(x=OR, y=model)) + 
             geom_errorbarh(aes(xmin=LCI, xmax=UCI), size=1, height=.1, color="#FFCC00") + 
             geom_point(size=2, color="#CC33FF") + 
             scale_y_discrete(labels = c("Adjusted", "Unadjusted")) +
             theme_bw() + 
             theme(panel.grid = element_blank()) + 
             labs(x="Model", y="OR (95% CI)") + 
             geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed") 

sor_p

# Export the plot
jpeg("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/plot_adjusted_unadjusted_regression_standardized_OR.jpeg")
sor_p
dev.off()
```

```{r}
# Plot AUROC of actual PRS in cases and controls together
jpeg("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/auc_standardizedcdprs_ukbb.jpeg")
rocobj <- plot.roc(ukb_cd$CD, ukb_cd$SCORE1_SUM,
                   main = "AUC-CD", 
                   percent=FALSE,
                   ci = TRUE,                  # compute AUC (of AUC by default)
                   print.auc = TRUE)
dev.off()
```

Plot ORs from non-standardized and standardized PRS in one plot from both models

```{r}
# Rbind results from all 4 models (standardized and non-standardized unadjusted and adjusted models)
all_or <- bind_rows(comb_or, scomb_or)
all_or$var[1:2] <- "PRS"

# Plot both ORs side by side
all_or_p <- ggplot(all_or, aes(x=OR, y=model)) + 
             geom_errorbarh(aes(xmin=LCI, xmax=UCI), size=1, height=.1, color="#FFCC00") + 
             geom_point(size=2, color="#CC33FF") + 
             geom_label_repel(aes(label=round(OR, digits = 2)), colour="black", segment.colour="black") + 
             scale_y_discrete(labels = c("Adjusted", "Unadjusted")) +
             facet_wrap(~var) + 
             theme_bw() + 
             theme(panel.grid = element_blank()) + 
             labs(x="OR (95% CI)", y="Model") + 
             geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed") 

all_or_p

# Export the plot
jpeg("~/Box Sync/Proj_GeRI_GI-irAEs/PRS_UKBB/Results/CD/10112022/prs_score/plot_adjusted_unadjusted_standardized_nonstandardized_OR.jpeg", width = 4450, height = 2800, res = 350)
all_or_p
dev.off()
```

